---
title: "ggplot"
author: "Zachary Noel"
date: "3/30/2021"
output:
  md_document:
    variant: markdown_github
  html_document: default
---

# Load in packages you need
```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
```


# Read in the data
```{r}
bull.richness <- readRDS("Bull_richness.RDS")
```

# Lets look at the data
```{r}
str(bull.richness)
```
These data contain data from a microbiome project where fungicides were applied to corn and soybean that have been managed under no-till or conventionally tillage. We sequenced the leaf microbiome before fungicides were applied, one week following fungicide applicaiton, and one month following fungicides. We also sequenced non-sprayed plots as controls. Then using the sequence data we counted the number of species of fungi in the Bulleribasidiaceae family.

In this dataset we are interested to know if the number of Bulleribasidiaceae species was reduced by the fungicide sprays, and how crop management influenced the disturbance. 

So we have the following variables to express on the plot: 
-Date Sampled: (i.e., before, one week, and one month after fungicide)
-Crop: Corn or soybean
-richness (i.e., number of species) - RESPONSE (Y) VARIABLE
-Fungicide: Control vs. Fungicide sprayed
-Treatment or crop management: Conventional, or No-till

# Set up a color blind friendly color pallete
```{r}
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

# Lets plot the data
we will start with soybean 
```{r}
bull.richness.soy <- bull.richness[bull.richness$Crop == "Soy",] # subset to soy data

ggplot(bull.richness.soy, aes(x = as.character(DateSampled), y = richness, color = Fungicide)) + 
  geom_boxplot() 
```
Woo! you made your first ggplot. But... its not quite what we want. Since it still doesn't show the different crop managements. And, well its kind of boring. 

Lets make two plots for different managements using facet
```{r}
ggplot(bull.richness.soy, aes(x = as.character(DateSampled), y = richness, color = Fungicide)) + 
  geom_boxplot() + 
  facet_wrap(~Treatment)
```

Lets give the plot custom x and y labels
we can do that using xlab() and ylab()
```{r}
ggplot(bull.richness.soy, aes(x = as.character(DateSampled), y = richness, color = Fungicide)) + 
  geom_boxplot() + 
  facet_wrap(~Treatment) + 
  xlab("") + 
  ylab("Bulleribasidiaceae richness") 
```

Maybe we want to actually show all of the datapoints to really show the distribution of the data. We can do that using geom_point(), which adds points to the plot, then we can tell it we want to "jitter" the points to make a distribution of the points over the boxplot. 
```{r}
ggplot(bull.richness.soy, aes(x = as.character(DateSampled), y = richness, color = Fungicide)) + 
  geom_boxplot() + 
  facet_wrap(~Treatment) + 
  xlab("") + 
  ylab("Bulleribasidiaceae richness") +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) 
```

Now, that's not half bad! Except this really is a time series, and I want to display the X axis as a nice date format. Also, I want to connect the time points with a line, while still showing the distribution of data. We can do that using the stat_summary function of ggplot. This is a very useful function, since it will plot the mean as the point, or bar, or line etc. Then you can also use it to plot the SE error bars. 

```{r}
ggplot(bull.richness.soy, aes(x = DateSampled, y = richness, color = Fungicide)) + 
  stat_summary(fun.y=mean,geom="line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  ylab("Bulleribasidiaceae \n richness") + 
  xlab("") + 
  facet_wrap(~Treatment) + 
  geom_jitter(width = 0.8, alpha = 0.5)
```

Great! Now let's actually test to see if this is significant. We can use the package ggpubr to put significance right on the plot. Of course, you would already have done an ANOVA to show that the interaction of crop management and fungicide treatment was significant... and test the assumptions of heteroskedasticity... but this is visual exploratory analysis.

```{r}
ggplot(bull.richness.soy, aes(x = DateSampled, y = richness, color = Fungicide)) + 
  stat_summary(fun.y=mean,geom="line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  ylab("Bulleribasidiaceae \n richness") + 
  xlab("") + 
  facet_wrap(~Treatment) + 
  geom_jitter(width = 0.8, alpha = 0.5) +
  stat_compare_means(method = "t.test", aes(label = ..p.signif..), hide.ns = TRUE, size = 6, label.y = 30) 
```

Now we've got something! lets make it pretty. With some colors from our color blind friendly pallete and change the legend text. Then I don't really like the boxes around the facets. So we can turn those white. Finally we can give it a title. I will save the plot as an object for displaying. 
```{r}
soy.bull <- ggplot(bull.richness.soy, aes(x = DateSampled, y = richness, color = Fungicide)) + 
  stat_summary(fun.y=mean,geom="line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  ylab("Bulleribasidiaceae \n richness") + 
  xlab("") + 
  facet_wrap(~Treatment) + 
  geom_jitter(width = 0.8, alpha = 0.5) +
  stat_compare_means(method = "t.test", aes(label = ..p.signif..), hide.ns = TRUE, size = 6, label.y = 30) +
  theme_classic() +
  scale_color_manual(values=cbbPalette, label = c("Control", "Fungicide"), name = "") +
  theme(
   strip.background = element_rect(color="white"), 
   strip.text.x = element_text(size = 12, color = "black")) +
  ggtitle("Soybean") 
soy.bull
```

Great! now lets apply this same plot to the corn. 
```{r}
bull.richness.corn <- bull.richness[bull.richness$Crop == "Corn",] # subset to corn data

corn.bull <- ggplot(bull.richness.corn, aes(x = DateSampled, y = richness, group = interaction(Fungicide, Treatment), color = Fungicide)) + 
  geom_jitter(width = 0.8, alpha = 0.5) +
  stat_summary(fun.y=mean,geom="line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  ylab("Bulleribasidiaceae \n richness") + 
  xlab("")+
  scale_color_manual(values=cbbPalette, label = c("Control", "Fungicide"), name = "") +
  stat_compare_means(method = "wilcox", aes(label = ..p.signif..), hide.ns = TRUE, size = 6, label.y = 30) +
  theme_classic() +
  facet_wrap(~Treatment, scales = "free_x") +
  theme(
   strip.background = element_rect(color="white", fill="white", size=1.5, linetype="solid"),
   strip.text.x = element_text(size = 12, color = "black"),
   legend.position="right",
   legend.title = element_blank()) + 
  ggtitle("Corn") 
corn.bull
```

Now we can put the two plots together into one using ggarage function in the ggpubr package
```{r}
ggarrange(corn.bull, soy.bull, common.legend = T, nrow = 1, ncol = 2, labels = c("A", "B"))
ggsave("Corn_Soy_Bull_Richness.pdf", dpi = 300) #save a pdf
```

ggarrange() has nice option to have a common legend between the two plots. This is useful in this case, so I turned that to "T" for true. You can also specify how you want the plots to be stuck together. I wanted them side-by-side so I put them in 1 row and 2 columns. Then I want a label as A and B for the two figures for the paper, so I can do that with the labels option. 

Next you can output the figure in several ways. You can ggsave() and specify the DPI, which is useful for publication. The resulting figure will be the same size as the panel figure. You can also specify the height and width in the function if you want. 

Alternatively, you can simply choose the export option on the figure panel and chose what you want to do. 




